Sexual selection and extinction: a macro-evolutionary approach

### 2018-06-27 Sexual Selection and Extinction.Rmd
27/6 Update

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
```

@@

To gain estimates for change in climate over the past 22,000 we can use the difference in annual mean temperature (BIO1) and annual precipitation (BIO12) between the LGM (CCSM4 models) and present values. Additionally we can calculate the difference in variability based on the standard deviation.

```{r}
#Change in mean temp
restricted.data$BIO1.LGM.Diff <- restricted.data$bioclim1 - restricted.data$cclgmbi1
#Change in variation temp
restricted.data$BIO1.LGM.var.Diff <- restricted.data$se.bioclim1 - restricted.data$se.cclgmbi1

@@

#Inspect these changes 
restricted.data %>% ggplot(aes(x=BIO1.LGM.Diff, y = BIO1.LGM.var.Diff))+
  geom_point()

restricted.data %>% ggplot(aes(x=BIO12.LGM.Diff, y = BIO12.LGM.var.Diff))+
  geom_point()
```

@@

##Preliminary Analysis of data

@@

```{r}
complete.dataframe %>% ggplot(aes(x=rab.ex.rate.lumped, y=rab.ex.rate.split))+
  geom_point()+
  scale_x_log10()+
  scale_y_log10()
```

The extinction rates of the lumped and split datasets show high similarity, insisting that we could use either.

Let us create a relatively simple model
```{r}
model.1 <- lm(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                data = restricted.data)
summary(model.1)
```

Now we can try a PGLS model 
```{r}
hackett.consensus.tree <- read.nexus('bamm_extinction/data/trees/hackett_dale_sub.consensus.nex')
restricted.data <- restricted.data %>% drop_na(rab.ex.rate.split)
rownames(restricted.data) <- restricted.data$TipLabel
pruned.tree<-drop.tip(hackett.consensus.tree,hackett.consensus.tree$tip.label[-match(restricted.data$TipLabel, hackett.consensus.tree$tip.label)])
pgls.model.1 <- gls(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                correlation = corBrownian(phy = pruned.tree), 
                data = restricted.data, 
                method = "REML")
summary(pgls.model.1)
```

Or perhaps OU model rather than Brownian:

@@

@@

```{r, eval=FALSE}
model_bayesian <- brm(
  rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res, data = restricted.data, 
  family = gaussian(), 
  cov_ranef = list(phylo = A),
                seed = 1,
                cores = 4, chains = 4, iter = 4000, #Run 4 chains in parallel for 4000 iterations (2000 are burn in)
                control = list(adapt_delta = 0.99, max_treedepth = 15))
```

```{r}
summary(model_bayesian)
```

Try looking at speciation rate using ES-Sim adapted by @Harvey_essim_2017 and found here https://github.com/mgharvey/ES-sim

```{r}
Essim.frame <- restricted.data$SDi
names(Essim.frame) <- restricted.data$TipLabel

essim.results <- essim(pruned.tree, Essim.frame, nsim = 1000, return.es = TRUE)
```

We can then add the essim.results and run a similar PGLS model
```{r}
restricted.data$ESsim <- essim.results[["es"]]

restricted.data %>% ggplot(aes(x=BIO12.LGM.Diff, y = (ESsim))) + 
  geom_point()+
  geom_smooth(method = "lm")
```

 Once we have our EsSim results we can use them in a linear model to assess the relationship between speciation and our predictors

```{r}
model.2 <- lm(ESsim ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                data = restricted.data)
summary(model.2)
```

Have a look at how these would look if we obtained model predictions for each data point.

```{r, fig.width=10, fig.height=5}
predict.frame <- cbind(predict(model.2),restricted.data)
predict.frame %>% ggplot(aes(y = `predict(model.2)`, x = SDi)) + 
  geom_point(aes(color = BLFamilyLatin))+
  scale_color_hue(h = c(0,100), l = 50) +
  geom_smooth(color = 'black')+
  theme_minimal()
```

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(gridExtra)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

@@

To gain estimates for change in climate over the past 22,000 we can use the difference in annual mean temperature (BIO1) and annual precipitation (BIO12) between the LGM (CCSM4 models) and present values. Additionally we can calculate the difference in variability based on the standard deviation, although these will be correlated and thus maybe using only one of them (either mean or variability difference) in the analysis may be better.

```{r, fig.width=7, fig.height=3.5}
#Change in mean temp
restricted.data$BIO1.LGM.Diff <- restricted.data$bioclim1 - restricted.data$cclgmbi1
#Change in variation temp
restricted.data$BIO1.LGM.var.Diff <- restricted.data$se.bioclim1 - restricted.data$se.cclgmbi1

@@

#Inspect these changes 
bio1.plot <- restricted.data %>% ggplot(aes(x=BIO1.LGM.Diff, y = BIO1.LGM.var.Diff))+
  geom_point()+
  theme_minimal()

bio12.plot <- restricted.data %>% ggplot(aes(x=BIO12.LGM.Diff, y = BIO12.LGM.var.Diff))+
  geom_point()+
  theme_minimal()

grid.arrange(bio1.plot, bio12.plot, ncol=2)
```

@@

##Preliminary analysis of data

###Examining BAMM rates

@@

```{r}
complete.dataframe %>% ggplot(aes(x=rab.ex.rate.lumped, y=rab.ex.rate.split))+
  geom_point()+
  scale_x_log10()+
  scale_y_log10()+
  theme_minimal()
```

###Fixed-effect models

The extinction rates of the lumped and split datasets show high similarity and we could arguably use either. Let's ude 'split':

First with a simple linear model that is not phylogenetically corrected:

```{r}
model.1 <- lm(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                data = restricted.data)
summary(model.1)
```

Secondly, we can use a PGLS model:

```{r}
hackett.consensus.tree <- read.nexus('bamm_extinction/data/trees/hackett_dale_sub.consensus.nex')
restricted.data <- restricted.data %>% drop_na(rab.ex.rate.split)
rownames(restricted.data) <- restricted.data$TipLabel
pruned.tree<-drop.tip(hackett.consensus.tree,hackett.consensus.tree$tip.label[-match(restricted.data$TipLabel, hackett.consensus.tree$tip.label)])
pgls.model.1 <- gls(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                correlation = corBrownian(phy = pruned.tree), 
                data = restricted.data, 
                method = "REML")
summary(pgls.model.1)
```

It is likely that given the underestimation of rate shifts by BAMM we obtain groups of similar speciation/extinction rates and thus when used in PGLS the highly correlated species will have near exact extinction/speciation rates. Giving less weight to these results.

Additionally, we can run a PGLS with an OU model rather than Brownian:

@@

###Bayesian models

@@

```{r, eval=FALSE}
# model_bayesian <- brm(
#   rab.ex.rate.split ~ SDi 
#                 + range.size.m2 
#                 + bioclim4
#                 + bioclim15
#                 + residuals.PC1
#                 + residuals.PC2
#                 + BIO1.LGM.Diff
#                 + BIO12.LGM.Diff
#                 + NPP.res, data = restricted.data, 
#   family = gaussian(), 
#   cov_ranef = list(phylo = A),
#                 seed = 1,
#                 cores = 4, chains = 4, iter = 4000, #Run 4 chains in parallel for 4000 iterations (2000 are burn in)
#                 control = list(adapt_delta = 0.99, max_treedepth = 15))
```

```{r}
#summary(model_bayesian)
```

###Diversification rate (DR) estimates as opposed to BAMM

Diversification rates can be simply estimated from thew 'tippiness of the phylogeny'. Here an equal splits (ES) value can be achieved that is weighted to more recent diversification. It was used by @Jetz_2012 and recently (in comparison with BAMM) in a study ivestigation bird diversity alongside elevation [@quintero_2018].WE CAN look at speciation rate using ES-Sim adapted by @Harvey_essim_2017 and found here https://github.com/mgharvey/ES-sim.

```{r, fig.width=8, fig.height = 4}
Essim.frame <- restricted.data$SDi
names(Essim.frame) <- restricted.data$TipLabel

essim.results <- essim(pruned.tree, Essim.frame, nsim = 1000, return.es = TRUE)
#Independent correlation results for ES-sim 
essim.stats <- cbind(essim.results[["rho"]], essim.results[["P Value"]])
colnames(essim.stats) <- c("rho", "P value")
essim.stats %>% pander(digits = 2)
```

These results suggest there is not a correlation between the diversification rate (log(ES)) and sexual dichromatism. However. We should look at this in a larger model and with phylogenetic correction. To do that we can obtain the ES values (tip rates) for each species, the log of these is the diversification rate used by @Jetz_2012 and @quintero_2018.

```{r}
restricted.data$ESsim <- essim.results[["es"]]

#OU model using ES-sim

pgls.model.4 <- gls(ESsim ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                correlation = corMartins(1, phy = pruned.tree), 
                data = restricted.data, 
                method = "REML")
summary(pgls.model.4)
```

We can now have a look at how these would look if we obtained model predictions for each data point.
```{r, fig.width=10, fig.height=5}
predict.frame <- cbind(predict(pgls.model.4),restricted.data)
predict.frame %>% ggplot(aes(y = `predict(pgls.model.4)`, x = SDi)) + 
  geom_point(aes(color = BLFamilyLatin))+
  scale_color_hue(h = c(0,100), l = 50) +
  geom_smooth(color = 'black')+
  theme_minimal()
```

###The sexual dimorphism dataset has an overdispersed distribution
